Convergence Studies of Mass Transport in Disks with Gravitational Instabilities. \\ I. The Constant Cooling Time Case

### 2011-11-09 Response_to_Referee1.txt

This is the response to the first referee report for ApJ submission MS#ApJ83382.

@@

My primary concern with the analysis is the neglect of the Reynolds stress component when calculating the
effective \alpha viscosity. The authors show that as the disk resolution is increased, more power is
transferred into higher m-structures approaching the dissipation scale in the code. It seems likely that the
shift also causes velocity correlations on this scale, which would produce extra angular momentum transport
and thus a higher effective \alpha.

* It should be noted that the measurement presented in Table 1 and Figure 3 of the manuscript are those of the
Fourier amplitudes, as defined in eqns 2 and 3. As can be seen by their definition, these values do not
directly measure the power in a given eignmode. In addition, each particular m value makes a relatively minor
contribution for large m (i.e. m > 20). The measurement of A_m also includes no phase information, so A_m is
not necessarily a good indicator of coherent density structures or organized velocity correlations.

@@

I do not understand the author's claim that it is too difficult or unreliable to measure Reynolds
stresses. This paper would be greatly improved by showing how the Reynolds term evolves with resolution. In
fact, I wonder whether all or some of the decrease in the transport at higher resolution would disappear if
this latter term were included.

* The difficulty arises from the fact that the Reynolds stress is typically defined as R_ij = rho*u'_i*u'_j
  where u'_i and u'_j are the fluctuations in the velocity field. The fluctuating pieces of the velocity field
  are defined as u'_i = u_i - u<avg>_i, but in a gravitationally unstable disk with extended non-axisymmetric
  spiral structures, the proper definition of u<avg> (particularly in the phi direction) is unclear. We
  measured the Reynolds stresses using several different definitions for u<avg> and obtained results that
  varied dramatically (by an order of magnitude in some cases). The cases we tried were: using an azimuthal
  average for u<avg>_phi and 0 for u<avg>_r, using an average +/- 2 cells in the direction of interest, using
  an equiangular average in the direction of interest (i.e. the number of cells used in the average for the
  phi direction increased with resolution), using a +/- 2 cell patch around the cell, and using an equiangular
  r,phi patch. As stated earlier, results for each of these cases varied substantially. It is also not clear
  what is the proper scale to average over, as one would like to measure the Reynolds stress caused by
  turbulent motion, not the ordered motion of spiral arms. In fact, when the Reynolds stress is measured by
  using a +/- 2 cell patch in both directions, the stresses measured are only a few percent of the
  gravitational torques.  (See attached figures.) response_to_referee1_fig1a.eps shows the measured torques
  from the Reynolds stress from 10 to 50 AU averaged from 12 to 18 ORP for LMAX = 64(black), 128(blue),
  256(green), and 512(red), while response_to_referee1_fig1b.eps shows the measured gravitational torques
  averaged over the same time period.

@@

In addition to measuring the Reynolds stresses directly, the authors could also look at the mass accretion
rate through rings in order to quantify the effective \alpha. The latter would be another way to get at the
contribution from Reynolds stresses.

* We were able to make this measurement and the mass accretion shows convergence similar to the gravitational
  torques. In addition we calculated an expected mass accretion based on the measured torques (see Boley et
  al. 2006 equation 22), and the gravitational torques alone are sufficient to account for the measured
  accretion rates. For LMAX = 64 the average over 10 to 40 AU of the absolute value of the measured radial
  mass flow rate was 8.9 x 10^-7 m_sun/yr; for LMAX = 128 it was 3.7 x 10^-7 m_sun/yr, for LMAX = 256 2.5 x
  10^-7 m_sun/yr, and for LMAX = 512 2.4 x 10^-7 m_sun/yr. The mass inflow rates are thus converging in a
  manner similar to the alpha measured from the gravitational torques alone. There is no indication of missing
  contributions to alpha from Reynolds stresses. This information is now included in Table 1 and in the text
  in Section 3.4.

@@

The authors could also measure the net heating in their simulation to quantify the effective \alpha via
dissipation. This would give again more insight into the effect of resolution on the critical cooling time or
the value of \alpha as a function of (Q). Even though the authors state that their results are converging and
not fragmenting, a lower effective \alpha should imply less heating, and thus increase the critical cooling
time. Of course, if \alpha is more and more dominated by Reynolds terms, than the heating rate and effective
alpha might not decline. This is a very important distinction. The comment in the summary that Q is increasing
with increasing resolution suggests that there is more heating, and thus more dissipation.

* Our code conserves angular momentum with more precision than it conserves energy, and so we have more
  confidence about changes in the radial position of mass elements due to changes in their specific angular
  momentum than in measurements of energy dissipation rates. The similar convergence of the measured 
  mass accretion rates and the gravitational alpha give us confidence that gravitational torques 
  dominate the mass transport.

@@

pg 6: A citation to Rice & Lodato 04-05 should be included as they tested the locality of transport and
dissipation.

* Citation included.

pg 6: A citation to Kratter et al 2010, Walsh et al 2010 would also be appropriate for models of embedded GI
unstable accretion disks.

* Kratter citation included, however we were unable to locate the Walsh paper referenced here.

@@

pg 7: I don't understand the emphasis on local vs "global" cooling time. I can see that a global cooling time
will make the entire disk go unstable at the same time (inner radii get to Q=1 faster), but it is no more
physical than the "local" choice, which allows for a cancellation in the energy equation and thus produces the
dimensionless criterion. The authors should provide some explanation of this issue, or of their choice. Of
course the difference between these two is irrelevant for Gammie's simulation.

* First paragraph on page 7 is rewritten to elucidate this issue.

@@

pg 8: "alpha" should be a letter

* Changed.

pg 9: "Source and flux terms..." The authors should explain what these are physically.

* Citations are now included where this is explained in detail.

@@

pg 20: "militate against" Do you mean mitigate?

* No, "militate" is the intended meaning.

pg 20:  It might be useful to include a footnote saying that you are using orbital periods instead of \Omega
to quote the critical values for the cooling time.

* Text modified as follows: "half the local orbital period (tcool~3/Omega)."

@@

pg 21: The sentence starting "For example, the Gammie 2001 simulations..." is confusingly written. Also, I
disagree with the description of the Meru and Bate results adding "weight to the claims by e.g. Boss..". It
seems clear that unless dissipation is dramatically non-local, there must be a relation between the rate of
dissipation and cooling. I don't think Boss is claiming that this doesn't exist, just that cooling isn't a
problem.

* Modified as follows: "For example, the Gammie (2001) simulations have revealed the most stringent cooling
  criterion of all the tcool-fragmentation studies, and due to the local approximation Gammie used, the
  simulations are also at very high resolution."

@@

fig 3 caption: "these quantities are roughly..." I don't know what this refers to

* Rewritten for clarity.

fig 5: While I appreciate the black and white legibility of this diagram, color might be useful for online
readers (or those who print in color) to distinguish between the various curves

* Both color and B/W versions of the figures have been provided.

### 2011-11-09 Paper_1.tex
Dick's changes to the referees report and paper, minor tweak to Figure 5.
	modified:   Paper_1.tex
	modified:   Response_to_Referee1.txt
	modified:   figures/Figure5.eps

The results above permit an assessment of appropriate azimuthal grids for the
problem under consideration and similar problems. The $l_{max} = 512$ simulation yields a value 
for $\alpha_{avg}$ of 0.020. This overestimates the extrapolated converged value of $\alpha_{avg}
= 0.017$ by $\sim 18$\%. The $l_{max} = 512$ simulation also overestimates the fractional
power in the high-order ($m = 2$ -- 7) modes by essentially the same percentage. However, it should be noted that the fractional difference between successive doubling of the resolution has decreased for $l_{max} = 512$. From $l_{max} =64$ to 128 the fractional difference in $\alpha_{avg}$ is 36\%. This decreases to 26\% for the jump from $l_{max} = 256$ to 512. Furthermore, the converged value represents infinite azimuthal resolution, which is, of course, impossible to achieve in a fixed grid scheme. 
%It is unclear as to what variation of $\alpha$ between two values of $l_{max}$ would represent convergence as $\alpha$ 
%itself is a dynamically varying quantity, see figure \ref{fig:alphavar}. 
In addition to the convergence of $\alpha_{avg}$, table \ref{tbl:ams} shows the convergence of the $\dot{M}$
from 10 -- 40 AU averaged from 12 to 18 ORP. The $\dot{M}_{avg}$ measurement shows convergence similar to that
of the $\alpha_{avg}$ measurement.

@@

\begin{table}
\begin{center}
\begin{tabular}{cccccc}
\multicolumn{5}{c}{$Q$, Fourier Amplitudes, and $\alpha$}\\ \hline\hline
$l_{max}$ & $Q_{avg}$   & $\langle A_{tot} \rangle$   & $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ &
$\alpha_{avg}$ & $\dot{M}_{avg} ~\mathrm{M_\odot/yr}$\\\hline
64   & 1.26 & 1.96 &  0.76 & 0.069 & 8.9$ \times 10^{-7}$ \\
128 & 1.24 & 1.91 &  0.62 & 0.044 & 3.7$ \times 10^{-7}$ \\
256 & 1.39 & 1.83 &  0.54 & 0.026 & 2.5$ \times 10^{-7}$ \\
512 & 1.48 & 1.95 &  0.49 & 0.020 & 2.4$ \times 10^{-7}$ \\ 
$\infty$ & & &0.43 & 0.017 & \\ \hline
\label{tbl:ams}
\end{tabular}
\caption{ All quantities are time-averaged over 12 -- 18 ORP.
$Q_{avg}$, $\alpha_{avg}$, and $\dot{M}_{avg}$  are also spatially
averaged over 10 -- 40 AU.  The entries for ``$\infty$'' are based on Richardson extrapolation,
as explained in the text.}  
\end{center}
\end{table}

>>>>

The results above permit an assessment of appropriate azimuthal grids for the problem under consideration and
similar problems. The $l_{max} = 512$ simulation yields a value for $\alpha_{avg}$ of 0.020. This
overestimates the extrapolated converged value of $\alpha_{avg} = 0.017$ by $\sim 18$\%. The $l_{max} = 512$
simulation also overestimates the fractional power in the high-order ($m = 2$ -- 7) modes by essentially the
same percentage. However, it should be noted that the fractional difference between successive doubling of the
resolution has decreased for $l_{max} = 512$. From $l_{max} =64$ to 128 the fractional difference in
$\alpha_{avg}$ is 36\%. This decreases to 26\% for the jump from $l_{max} = 256$ to 512. Furthermore, the
converged value represents infinite azimuthal resolution, which is, of course, impossible to achieve in a
fixed grid scheme.
%It is unclear as to what variation of $\alpha$ between two values of $l_{max}$ would represent convergence as $\alpha$ 
%itself is a dynamically varying quantity, see figure \ref{fig:alphavar}.

%DICK: I made changes here.
In addition to the convergence of $\alpha_{avg}$, table 1 illustrates the convergence of the measured net
inward mass transport rates $\dot{M}$ over the time interval 12 -- 18 ORP. We tabulate the value of
$|\dot{M}|$ spatially averaged over 10 -- 40 AU. The absolute value of the net $\dot{M}$ at each $r$ is taken
prior to the spatial average because $\dot{M}$ varies in sign with $r$. It is the magnitude of the typical net
mass transport rates that we wish to compare. The $|\dot{M}|_{avg}$ measurements show convergence similar to
that of the $\alpha_{avg}$ measurement based on the gravitational torques. This gives us confidence that no
important contributions from Reynolds stresses are being overlooked. Moreover, we have used eq. (22) of
\citet{boley2006} to estimate mass inflow rates based on the measured gravitational torques and find that
these torques are indeed sufficient to account for the measured $\dot{M}$.

@@

%DICK: I made changes here.
\begin{table}
\begin{center}
\begin{tabular}{cccccc}
\multicolumn{5}{c}{$Q$, Fourier Amplitudes, $\alpha$, and $\dot{M}$}\\ \hline\hline
$l_{max}$ & $Q_{avg}$   & $\langle A_{tot} \rangle$   & $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ &
$\alpha_{avg}$ & $|\dot{M}|_{avg} ~\mathrm{M_\odot/yr}$\\\hline
64   & 1.26 & 1.96 &  0.76 & 0.069 & 8.9$ \times 10^{-7}$ \\
128 & 1.24 & 1.91 &  0.62 & 0.044 & 3.7$ \times 10^{-7}$ \\
256 & 1.39 & 1.83 &  0.54 & 0.026 & 2.5$ \times 10^{-7}$ \\
512 & 1.48 & 1.95 &  0.49 & 0.020 & 2.4$ \times 10^{-7}$ \\ 
$\infty$ & & &0.43 & 0.017 & \\ \hline
\label{tbl:ams}
\end{tabular}
\caption{ All quantities are time-averaged over 12 -- 18 ORP.
$Q_{avg}$, $\alpha_{avg}$, and $|\dot{M}|_{avg}$  are also spatially
averaged over 10 -- 40 AU.  The entries for ``$\infty$'' are based on Richardson extrapolation,
as explained in the text.}  
\end{center}
\end{table}
